---
title: "assignment_2"
author: "Matta Nguyen"
date: "2025-02-14"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part 1: 
### Part 1a

(a) The first prior formulation was
```θj |α,β∼ Beta(α,β)
α,β∼ iid Expon(0.001)
```

  (i) [2 pts] Independently simulate 1000 pairs (α,β) from their hyperprior, and produce a scatter plot of log(β) versus log(α).


```{r Similation}
set.seed(123)

nsims <- 1000

# Stimulate alpha and beta
alpha_a <- rexp(nsims, rate = 0.001)
beta_a <- rexp(nsims, rate = 0.001)
```

```{r Scatter Plot}
plot(log(alpha_a), log(beta_a),
     xlab="log(α)",
     ylab="log(β)",
     main="1a.Scatterplot of log(β) vs log(α)",
     pch=20)
```

(ii) [2 pts] Using the simulated pairs (α,β), forward-simulate θ1, and produce a
histogram of the result (an approximation of its marginal prior)

```{r}
theta_1_a <- rbeta(1000, alpha, beta)
```


```{r Histogram}

hist(theta_1_a,
     breaks=30,
     main="Histogram of Simulated θ1|α,β Values",
     xlab="θ1|α,β",
     freq=FALSE)

```

### Part 1b

(b) The second prior formulation was
θj |α,β∼ Beta(α,β)
α = φ1/φ2
2 β = (1−φ1)/φ2
2
φ1 ∼ U(0,1) φ2 ∼ Expon(0.001)

(i) (ii) [2 pts] Independently simulate 1000 pairs (α,β) from their hyperprior, and
produce a scatterplot of log(β) versus log(α).

```{r}
# Stimulate phi1 and phi2 based on formula

phi1 <- runif(nsims, 0,1) # phi1 with uniform (0,2)
phi2 <- rexp(nsims, rate = 0.001) # phi2 with Expon(0.001)

# Calculate α,β

alpha_b <- phi1/phi2^2
beta_b <- (1-phi1)/phi2^2

```

Scaterplot 

```{r scatterplot}
plot(log(alpha_b), log(beta_b),
     xlab="log(α)",
     ylab="log(β)",
     main="1b. Scatterplot of log(β) vs log(α)",
     pch=20)
```

[2 pts] Using the simulated pairs (α,β), forward-simulate θ1, and produce a
histogram of the result (an approximation of its marginal prior).

```{r}
theta_1_b <- rbeta(nsims, alpha_b, beta_b)

# histogram:

hist(theta_1_b,
     breaks=30,
     main="Histogram of Simulated θ1_b|α,β Values",
     xlab="θ1_b|α,β",
     freq=FALSE)
```

## Part 2
### Part 2a:

According to the provided Bayesian hierarchical model we have:

ψ0 ∼ N (0,1000^2) 
-> Because the variance of this is too big. The prior could be considered as **flat prior**. Therefore the density is **p(ψ₀) ∝ 1 **

σ0 ∼ U(0,1000) approximates this **improper prior** . Therefore density would be **σ₀: p(σ₀) ∝ 1/σ₀ for σ₀ > 0 **

### Part 2b: Draw DAQ



### Part 2c: JAGS Nodel Statement

Utilzing asgn2template.bug

```{r JAGs}
model {
  for (j in 1:12) {
    psihat[j] ~ dnorm(psi[j],1/(sigma[j]*sigma[j]))
    psi[j] ~ dnorm(psi0, 1/sigmasq0)
  }

  psi0 ~ dnorm(0,1/1000^2)
  sigma0 ~ dunif(0,1000)

  sigmasq0 <- pow(sigma0,2) # calculate σ₀²
}
```

### Part 2d: 

Data Set up

```{r}
data_raw <- read.table("thenumbers.txt")

# Transform data in thenumbers.txt to vector
psihat <- c(data_raw[,2], data_raw[,5], data_raw[,8])
sigma <- c(data_raw[,3], data_raw[,6], data_raw[,9])

# Create datalist
data_list <- list (
  psihat = psihat,
  sigma = sigma
)

# Testing and showing the datalist. 
head(data_list)
length(data_list$psihat) #should be 12
length(data_list$sigma) #should be 12
```
Setting up JAGS model

```{r}
library(rjags)
jags.model.default <- jags.model("asgn2template.bug", data_list, n.chains = 3)
```
### Part 2e:

Running burn-in and draw sample with 10k to 100k iterations

```{r}
# Burn-in
update(jags.model.default,10000)

# Draw sample
samples <- coda.samples(jags.model.default, 
                       variable.names = c("psi0", "sigmasq0"), 
                       n.iter = 100000)

# Summary 
summary_stats <- summary(samples)
print(summary_stats)
```
For both ψ0 and σ20 (not σ0), produce a posterior  and also graphical estimates of the posterior densities. Explicitly give the approximations of
their posterior expected values, posterior standard deviations, and 95% central
posterior intervals. 

Numerical Summary

```{r}
# Extract posterior means and SDs
post_mean_psi0 <- summary_stats$statistics["psi0", "Mean"]
post_sd_psi0 <- summary_stats$statistics["psi0", "SD"]
post_mean_sigmasq0 <- summary_stats$statistics["sigmasq0", "Mean"]
post_sd_sigmasq0 <- summary_stats$statistics["sigmasq0", "SD"]

# Extract 95% credible intervals
cred_intervals <- summary_stats$quantiles[, c("2.5%", "97.5%")]

# Displaying result:
cat("\nPosterior summary for ψ₀:\n")
cat("Mean:", round(post_mean_psi0, 3), "\n")
cat("SD:", round(post_sd_psi0, 3), "\n")
cat("95% CI: [", round(cred_intervals["psi0","2.5%"], 3), 
    ",", round(cred_intervals["psi0","97.5%"], 3), "]\n")

cat("\nPosterior summary for σ₀²:\n")
cat("Mean:", round(post_mean_sigmasq0, 3), "\n")
cat("SD:", round(post_sd_sigmasq0, 3), "\n")
cat("95% CI: [", round(cred_intervals["sigmasq0","2.5%"], 3), 
    ",", round(cred_intervals["sigmasq0","97.5%"], 3), "]\n")
```

Graphical estimate

```{r}
plot(density(as.matrix(samples)[,"psi0"]), 
     main="Posterior density of ψ₀",
     xlab="ψ₀")

plot(density(as.matrix(samples)[,"sigmasq0"]), 
     main="Posterior density of σ₀²",
     xlab="σ₀²")

```

### Part 2f:

a new case-control study is to be performed, and assume that its log-odds
standard error (new σ) will be 0.125.

**(i) New DAG**

**(ii) Adjusted JAG**

```{r}
new_model <- "
  model {
    # Original model
    for (j in 1:12) {
    psihat[j] ~ dnorm(psi[j],1/(sigma[j]*sigma[j]))
    psi[j] ~ dnorm(psi0, 1/sigmasq0)
  }

  psi0 ~ dnorm(0,1/1000^2)
  sigma0 ~ dunif(0,1000)

  sigmasq0 <- pow(sigma0,2) # calculate σ₀²
  
  # New prediction
  psi.new ~ dnorm(psi0, 1/sigmasq0)
  psihat.new ~ dnorm(psi.new, 1/(new.sigma*new.sigma))
  
  # Indicator for significance
  significant <- step(psihat.new - 2*new.sigma)
  }"
```

Update Data

```{r}
# Update data list with new sigma
data_list_pred <- c(data_list, list(new.sigma = 0.125))
```

Fit Model

```{r}
new.jags.model <- jags.model(textConnection(new_model), 
                             data = data_list_pred, 
                             n.chains = 3)
```
Burn in & Sample

```{r}
update(new.jags.model, 10000)

# Draw samples
samples_pred <- coda.samples(new.jags.model, 
                           variable.names = c("psihat.new", "significant"), 
                           n.iter = 100000)

summary_pred <- summary(samples_pred)
```

**(iii) Posterior Mean & Postetior Standard Deviation**

```{r}
summary_pred <- summary(samples_pred)

cat("\nPredictive distribution for new study's estimated log-odds ratio:\n")
cat("Mean:", round(summary_pred$statistics["psihat.new", "Mean"], 3), "\n")
cat("SD:", round(summary_pred$statistics["psihat.new", "SD"], 3), "\n")
cat("95% Predictive Interval: [",
    round(summary_pred$quantiles["psihat.new", "2.5%"], 3), ",",
    round(summary_pred$quantiles["psihat.new", "97.5%"], 3), "]\n")
```
**(iv) Probability of significant**

```{r}
prob_significant <- mean(as.matrix(samples_pred)[, "significant"])
cat("\nProbability of finding significant positive result:",
    round(prob_significant, 3), "\n")
```